Supplementary Information

### 2020-12-22 doc/manuscript/MEE_manuscript.Rmd
results

###Windows selected: 
# BCI: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P cApr-cJul(4mo), TMXpJul-cJul(13mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCI and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCI, 9 mo at HKK) also included the majority (BCI) or all (HKK) of the dry season.
At three extratropical sites (LDW, NIO, and CB), the most influential windows were limited to the current spring and early summer, whereas optimal windows at two sites extended back to the previous fall (LT) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

*rearranged to here.*

The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig. `r clim_sens`).

Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`): wet seasons in the tropics (precipitation variables at both BCI and HKK; $T_{max}$ at HKK) or late spring/ early summer in the extratropics (n= 5 of 8 each for temperature and precipitation variables).
In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
There were also cases where previous growing season conditions were influential, including  temperature variables at Niobrara and Cedar Breaks. 
In two cases ($PPT$ at SCBI and $T_{max}$ at Cedar Breaks), the window of influence spanned >12 months to include past and current growing seasons.

(*Results up to this point are current as of Dec. 21. Those following need to be updated. I also want to reorganize to integrate with the content above*)

Temperature responses were included in the best model at all sites and *for the majority of species* (Fig. `r clim_sens`).
Specifically, a first-order temperature term was included in the best model for # of # site-species combinations, a second-order polynomial term significant for #.
The majority (#%) of significant temperature responses were negative.
Exceptions included positive responses of
*Tsuga canadensis* (HF) to March $PET$,
all three species at Zofin to to January-March $T_{max}$,
**NE TBD**,
and *Picea mariana* at Scotty Creek. 
We note, however, that the temperature response of the latter is negative above the long-term mean temperature when the analysis is split pre- and post-1970 (Figs. **ADD SI FIGS**).

@@

**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
While $RW$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
Similar patterns of approximately exponential decline in $RW$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig. `r size`).
At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.

>>>>

###Windows selected: 
# BCI: precip dry season (12mo), T dry season (2 mo) 
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
#LDW: precip c June (1 mo), PET current June (1mo)
#HF: PDF pMay (1mo), PET cMar (1mo)
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
#NIO: wet cMay (1mo), TMP pJune (1mo)
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
#CB: P cApr-cJul(4mo), TMXpJul-cJul(13mo)
#SC: P pJun-pAug(3mo), TMXcJul (1mo)
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCI and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCI, 9 mo at HKK) also included the majority (BCI) or all (HKK) of the dry season.
At three extratropical sites (LDW, NIO, and CB), the most influential windows were limited to the current spring and early summer, whereas optimal windows at two sites extended back to the previous fall (LT) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`). 
At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites. 
$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig. `r clim_sens`).
Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCI), late winter/early spring (HF, ZOF), the previous growing season (NIO), or a 13-month window extending from past to current growing season (CB).
Within the tropics, there was minimal effect of temperature at BCI and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. **CREATE AN SI FIGURE!**).

@@

**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
While $RW$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
On one end of the spectrum, the light-demanding species *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
Similar patterns of approximately exponential decline in $RW$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig. `r size`).
At the other end of the spectrum, a number of shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.
